---
title: "Assignment 1"
echo: TRUE
message: FALSE
warning: FALSE
---

* Each attendee is requested at **minimum** to present some analysis of:
  * **their** countries catch statistics as reported to FAO
  * a selected species of interest accross all the countries as reported to FAO
* If attendee want rather to present an analysis done during this meeting on their own data they are more than welcome to do so.

# Action

Generate an RStudio-project. Get this data on the computer **now** by running this code:

```{r}
library(tidyverse)
fao <- read_csv("https://heima.hafro.is/~einarhj/data/fao-capture-statistics_area31.csv")
fao |> write_rds("fao.rds")
```

Make sure that you know where you actually stored the "fao.rds" on your computer (inside your Rproject-directory, but you have to know where that is). You can read the data in at any stage again from your own computer by this code:

```{r}
fao <- read_rds("fao.rds")
```

# Background information on the FAO data

The fao dataset contains capture catch statistics from FAO area 31 by species and country for the years 1950 to 2021 (66673 records). The data set contains the following variables:

-   **area**: FAO area code (only area 31)
-   **year**: year
-   **country**: country name in english
-   **species**: "species" name in english
-   **catch**: catch volume (kg if measure is 'Q_ltw')
-   **measure**: unit of measure, mostly 'Q_ltw'
-   **sid**: 3 letter species code
-   **latin**: latin name of "species"
-   **country_iso_3**: 3 letter country code

# Suggested analysis

* There is no specific suggestion except we like some plots and summary tables.

# Product

* The product can be an R-script that includes comments about the approach used (use # at the beginning of a line)
* To create a html-presentation from the R-script that you can present do:
  * File -> Compile Report ...
* The objective is not only presenting your code and results (graphs/tables) but communicate what information/inference you make based on the analysis.
* Send the R-script and the html-file to Einar and he will put this all analysis as an anonymous document on the course web-site.

# Deadline

* Those presenting data analysis on the FAO data will give presentation on 2023-06-24 11:30:00
  * Deadline for sending the R-code and html-file to Einar: 2023-06-24 08:30:00
  * Expect that additional assignments will be made
* Those presenting data analysis on their own data will give a presentation on 2023-06-26, time to be determined.
  * No additional assignment expected
