---
title: "Bits and pieces"
echo: FALSE
message: FALSE
warning: FALSE
---

# On dataframes and vectors

So far we have been dealing with data-frames (e.g. the minke data):

```{r}
library(tidyverse)
w <- read_csv("ftp://ftp.hafro.is/pub/data/csv/minke.csv")
```

We can check the class of the object w via:

```{r}
class(w)    # key thing for now is that we see this is a "data.frame"
```

We access the values of each of the variables by using the '$'-sign:

```{r}
w$length
w$weight
```

We can even access individual or set of values using the '[ ]'

```{r}
w$length[1]              # length of the first observations
w$weight[2]              # weight of the second observation
w$maturity[1:10]         # maturity of the first 10 observations
```

Let's for now isolate the length measurements and check the class:

```{r}
lengths <- w$length
class(lengths)
```

This tells us that the object "lengths" is a numeric vector. 


# On functions and arguments

Almost all the stuff you do in R is via some function call. All functions of course have a specific name and most functions have arguements. A typical structure of a function looks something like this:

```{r eval = FALSE}
function_name(argument1 = value1, argument2 = value2, ...)
```




# ...