---
title: "Background on the data used"
---

# Fish dataset

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(omar)
con <- connect_mar()
set.seed(314)
d <- 
  omar::ch_station(con) |> 
  left_join(omar::ch_sample(con)) |> 
  inner_join(omar::bi_age(con)) |> 
  filter(sclass %in% c(30, 35),
         sid %in% c(1, 2, 3, 9),
         !is.na(age),
         age <= 14) |> 
  collect(n = Inf) |> 
  mutate(lclass = cut(length, c(0, 30, 40, 50, 60, 70, 80, Inf), include.lowest = TRUE)) |> 
  group_by(sid, lclass) |> 
  sample_n(20) |> 
  ungroup() |> 
  select(-lclass)
d |> ggplot(aes(wt, gwt, colour = factor(sid))) + geom_point()
d |> ggplot(aes(liver, gonads, colour = factor(sid))) + geom_point()
d |> ggplot(aes(lon, lat, colour = factor(sid))) + geom_point()
d |> ggplot(aes(length, gwt, colour = factor(sid))) + geom_point()
d |> ggplot(aes(age, length, colour = factor(sid))) + geom_jitter() + geom_smooth()
d |> ggplot(aes(age, wt, colour = factor(sid))) + geom_jitter() + geom_smooth()

```


# Minke dataset

Location: ftp://ftp.hafro.is/pub/data/csv/minke.csv

The minke whale dataset contains biological measurements from 192 scientific catches of minke whales between the year 2003 and 2007. The data set contains the following fields:

-   **whale.id**: Unique identifier for the whale
-   **date.caught**: the date when the whales was caught
-   **lat**: latitude
-   **lon**: longitude
-   **area**: Derived from location (North/South)
-   **length**: length of the whale (cm)
-   **weight**: weight of the whale (kg)
-   **age**: age of the whale (years)
-   **sex**: Male or Female
-   **maturity**: maturity status of the whale
-   **stomach.volume**: volume of the stomach content (liters)
-   **stomach.weight**: weight of the stomach content (kg)
-   **year**: the year when the whale was caught

You can import this data directly into R by:

```{r message = FALSE}
library(tidyverse)
d <- read_csv("ftp://ftp.hafro.is/pub/data/csv/minke.csv")
```


