---
title: "Background on the data used"
---

## FAO capture statistics in area 31

Location: <https://heima.hafro.is/~einarhj/data/fao-capture-statistics_area31.csv>

The fao dataset contains capture catch statistics from FAO area 31 by species and country for the years 1950 to 2021 (66673 records). The data set contains the following variables:

-   **area**: FAO area code (only area 31)
-   **year**: year
-   **country**: country name in english
-   **species**: "species" name in english
-   **catch**: catch volume (kg if measure is 'Q_ltw')
-   **measure**: unit of measure, mostly 'Q_ltw'
-   **sid**: 3 letter species code
-   **latin**: latin name of "species"
-   **country_iso_3**: 3 letter country code

You can import this data directly into R by:

```{r message = FALSE}
library(tidyverse)
fao <- read_csv("https://heima.hafro.is/~einarhj/data/fao-capture-statistics_area31.csv")
```

## Minke dataset {#minke-dataset}

Location: <ftp://ftp.hafro.is/pub/data/csv/minke.csv>

The minke whale dataset contains biological measurements from 192 scientific catches of minke whales between the year 2003 and 2007. The data set contains the following variables:

-   **whale.id**: Unique identifier for the whale
-   **date.caught**: the date when the whales was caught
-   **lat**: latitude
-   **lon**: longitude
-   **area**: Derived from location (North/South)
-   **length**: length of the whale (cm)
-   **weight**: weight of the whale (kg)
-   **age**: age of the whale (years)
-   **sex**: Male or Female
-   **maturity**: maturity status of the whale
-   **stomach.volume**: volume of the stomach content (liters)
-   **stomach.weight**: weight of the stomach content (kg)
-   **year**: the year when the whale was caught

You can import this data directly into R by:

```{r message = FALSE}
library(tidyverse)
minke <- read_csv("ftp://ftp.hafro.is/pub/data/csv/minke.csv")
```

## Appendix

### The code to get the FAO data

```{r, eval = FALSE}
library(tidyverse)
# Check if this can be done automatically - i.e. find the latest version
pth <- "https://www.fao.org/fishery/static/Data/Capture_2023.1.1.zip"
fil <- basename(pth)
download.file(pth, destfile = paste0("data-raw/", fil))
unzip(paste0("data-raw/", fil), exdir = "data-raw")
cntr <-
  read_csv("data-raw/CL_FI_COUNTRY_GROUPS.csv") |>
  janitor::clean_names() |>
  select(cid = un_code, country_iso3 = iso3_code, country = name_en)
species <-
  read_csv("data-raw/CL_FI_SPECIES_GROUPS.csv") |>
  janitor::clean_names() |>
  select(sid = x3a_code, species = name_en, latin = scientific_name)
d <-
  read_csv("data-raw/Capture_Quantity.csv") |>
  janitor::clean_names() |>
  filter(area_code == "31") |>
  select(cid = country_un_code,
         sid = species_alpha_3_code,
         area = area_code,
         year = period,
         catch = value,
         measure) |>
  left_join(cntr) |>
  left_join(species) |> 
  select(area, year, country, species, catch, measure, sid, latin, country_iso3)
d |> write_csv("/net/www/export/home/hafri/einarhj/public_html/data/fao-capture-statistics_area31.csv")
system("chmod -R a+rX /net/www/export/home/hafri/einarhj/public_html/data/")
```

```{r}
geo::island |> 
  write_csv("/net/www/export/home/hafri/einarhj/public_html/data/island.csv")
system("chmod -R a+rX /net/www/export/home/hafri/einarhj/public_html/data/")
```

