---
title: "Import"
message: FALSE
warning: FALSE
---

## ...

## Importing directly from tip-files

We can read tip-files directly into R using the function `foreign::read.dbf`. I you have not installed the {foreign}-package before you have to install it first by running:

```{r eval = FALSE}
install.packages("foreign")
```

Then do:

```{r}
library(tidyverse)
library(foreign)
```

If you have a tip-file (the suffix is .DBF) you could try something like this:

```{r, eval = FALSE}
read.dbf("your/path/to/the/data/tip11.DBF", as.is = TRUE)
```

Of course you need to replace "your/path/to/the/data/tip11.DBF" with your path of your data on your own computer. If you do not have tip file, but want to give it a go run this code:

```{r}
# download some species lookup-TIP data taken from some anonymous country
download.file(url = "https://heima.hafro.is/~einarhj/data/FISHCODE.DBF", destfile = "FISHCODE.DBF", mode = "wb")
fishcode <- read.dbf("FISHCODE.DBF", as.is = TRUE)
glimpse(fishcode)
```

